\emph{icepack}: a new glacier flow modeling package in Python, version 1.0

### 2020-12-12 icepack.tex
Final corrections

@@

@@

Many aspects of glacier physics are not completely understood.
For example, the most commonly used sliding law is the power law
\begin{equation}
    \tau_b = -C|u|^{\frac{1}{m} - 1}u
    \label{eq:weertman-sliding}
\end{equation}
for some exponent $m$.
Older research assumed $m = 3$ based on the theory of regelation \citep{weertman1957sliding}, which has since been referred to as the Weertman sliding law.
When $m = \infty$, the basal shear stress is independing of the sliding speed; this is referred to as perfect plasticity.
More recently, \citet{schoof2005effect} proposed the functional form
\begin{equation}
    \tau_b = -C\left(\frac{|u|}{|u| + u_c}\right)^{\frac{1}{m}}\frac{u}{|u|}
\end{equation}
where $u_c$ is some critical speed.
This sliding law acts like the Weertman law when $|u| \ll u_c$ and like the plastic law when $|u| \gg u_c$.
This latter equation has been found to agree best with laboratory experiments \citep{zoet2020slip}.

@@

One of our goals with icepack is to facilitate experimentation with the model physics, even for novice users.
Of the programming languages that are commonly used for scientific computing, only Python and possibly Julia would appear meet these needs.
To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress, which Python accomodates easily.
For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function, which is undesirable.
In C++, one could avoid changing the signature of the diagnostic solve routine by (1) using variadic templates, (2) wrapping the inputs in a class, or (3) passing all arguments in a dictionary.
Using variadic templates or wrapping the inputs in a class would require users to know more about generic or object-oriented programming than a novice might.
Using a dictionary data structure to pass arguments is relatively easier but would be more idiomatic in Python than in C++.

In icepack, users can substitute any diagnostic model component for the parameterization of their choosing, including adding new fields.
From the user's perspective, substituting model physics components does not require any advanced language features beyond keyword arguments.
To understand how this is possible, we'll briefly describe the path that the input fields take through the program.
\begin{enumerate}
    \item The user passes all arguments to the diagnostic solve procedure by keyword, essentially in a dictionary mapping names to fields.
    \item The diagnostic solve procedure creates a symbolic representation of the action functional by summing up several terms, like the viscosity, basal friction, etc.
        Calculating these terms is delegated to specialized procedures for each term.
        Each term procedure gets the entire collection of fields.
    \item The routine that calculates the terms of the action selects which fields it actually needs from the argument dictionary, and then creates the symbolic representation of that term.
    \item Finally, once the symbolic representation of the action functional has been created, all the responsibility passes to the nonlinear solver.
\end{enumerate}
To substitute different model components, the user intervenes at step 3.
Each model object -- shallow ice, shallow stream, etc. -- is initialized with a default set of routines to calculate the terms of that model's action functional.
The user can replace these default routines with one of their own choosing by passing the function of their choice when that model object is initialized.

@@

@@

Using variational principles to express all constitutive laws is less flexible than specifying the sliding law directly and this is a distinct disadvantage.
We were able to implement a sliding law that exhibits many of the characteristics of the Schoof sliding law.
\citet{asay2016experimental} also suggest using a sliding law that transitions sharply to exact plasticity above the critical speed.
Expressing this sliding law in UFL requires a conditional in the velocity and is thus no longer differentiable, causing the forward solver to crash.
The numerical advantages of using variational principles are so great, however, that we view this tradeoff as acceptable.

>>>>

We verified the correctness of our implementation by check that numerical results converge at the expected rate to the analytical Beuler profile for a circular symmetric ice sheet \citep{greve2009dynamics}.

@@

We verified the correctness of the ice shelf model by checking that numerical results converge at the expected rate to an analytical solution for the velocity with a linearly sloping thickness in a rectangular domain \citep{greve2009dynamics}.
When there is basal friction, the shallow stream equations do not have a simple analytical solution.
To get an exact solution, we chose the ice velocity and thickness a priori and used the computer algebra system SymPy \citep{sympy} to generate a friction coefficient that makes these fields an exact solution.
We then checked that numerical results converge to this manufactured solution at the expected rate.

@@

Many aspects of glacier physics are not completely understood.
For example, the most commonly used sliding law is the power law
\begin{equation}
    \tau_b = -C|u|^{\frac{1}{m} - 1}u
    \label{eq:weertman-sliding}
\end{equation}
for some exponent $m$.
Older research assumed $m = 3$ based on the theory of regelation \citep{weertman1957sliding}, which has since been referred to as the Weertman sliding law.
When $m = \infty$, the basal shear stress is independing of the sliding speed; this is referred to as perfect plasticity.
More recently, \citet{schoof2005effect} proposed an alternative form that acts like the Weertman law when $|u| \ll u_c$ and like the plastic law when $|u| \gg u_c$.
A simplified form of this model can be expressed as
\begin{equation}
    \tau_b = -C\left(\frac{|u|}{|u| + u_c}\right)^{\frac{1}{m}}\frac{u}{|u|}
\end{equation}
where $u_c$ is some critical speed.
This latter equation has been found to agree best with laboratory experiments on till \citep{zoet2020slip} and in reproducing observed velocity variations \citep{joughin2019regularized}.

@@

One of our goals with icepack is to facilitate experimentation with the model physics, even for novice users.
Of the programming languages that are commonly used for scientific computing, only Python and possibly Julia would appear to meet these needs.
To support use cases like implementing the Schoof sliding law, it must be possible not just to change the value of a single parameter but to completely alter the functional form of a given model component.
For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model component.
In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress, which Python accomodates easily.
For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function, which is undesirable.
In C++, one could avoid changing the signature of the diagnostic solve routine by (1) using variadic templates, (2) wrapping the inputs in a class, or (3) passing all arguments in a dictionary.
Using variadic templates or wrapping the inputs in a class would require users to know more about generic or object-oriented programming than a novice might.
Using a dictionary data structure to pass arguments is relatively easier but would be more idiomatic in Python than in C++.

In icepack, users can substitute any diagnostic model component for the parameterization of their choosing, including adding new fields.
From the user's perspective, substituting model physics components does not require any advanced language features beyond keyword arguments.
To understand how this is possible, we'll briefly describe the path that the input fields take through the program.
\begin{enumerate}
    \item The user passes all arguments to the diagnostic solve procedure by keyword.
    \item The diagnostic solve procedure creates a symbolic representation of the action functional by summing up several terms, like the viscosity, basal friction, etc.
        Calculating these terms is delegated to specialized procedures for each term.
        Each term procedure gets the entire collection of fields.
    \item The routine that calculates the terms of the action selects which fields it actually needs from the argument dictionary, and then creates the symbolic representation of that term.
    \item Finally, once the symbolic representation of the action functional has been created, all the responsibility passes to the nonlinear solver.
\end{enumerate}
To substitute different model components, the user intervenes at step 3.
Each model object -- shallow ice, shallow stream, etc. -- is initialized with a default set of routines to calculate the terms of that model's action functional.
The user can replace these default routines with one of their own choosing by passing the function of their choice when that model object is initialized.

@@

\subsection{Performance}

Icepack largely inherits the performance capabilities of the Firedrake package, for which we refer to the benchmarks in \citet{rathgeber2016firedrake}.
Firedrake is in turn built on PETSc, which includes a rich suite of nonlinear solvers and preconditioners that have been demonstrated to scale up to hundreds of processors \citep{balay2019petsc}.
Firedrake also includes special features for defining sophisticated solvers and preconditioners that take advantage of problem-specific structure \citep{kirby2018solver}.
Icepack exposes these tools through the interface of the flow solver objects, so users can select any solver and preconditioner from PETSc.

We have mainly developed icepack for process-scale studies of individual glaciers or drainage basins.
For the demonstrations presented below, nearly all simulations run in a matter of minutes to hours on a single core.
Larger problems, such as continental-scale modeling, will require more sophisticated and possibly problem-specific approaches.
For example, a rudimentary preconditioner for the hybrid model system could use the degree-0 model as the coarse space in a multigrid-type approach.
These optimizations will be the subject of future work.

@@

Using variational principles to express all constitutive laws is less flexible than specifying the sliding law directly and this is a distinct disadvantage.
We were nonetheless able to implement a sliding law that exhibits the important characteristics of the Schoof sliding law.
\citet{asay2016experimental} also suggest using a sliding law that transitions sharply to exact plasticity above the critical speed.
Expressing this sliding law in UFL requires a conditional in the velocity and is thus no longer differentiable, causing the forward solver to crash.
The numerical advantages of using variational principles are so great, however, that we view this tradeoff as acceptable.
Moreover, there are few physical reasons to prefer one form over another, as long as the asymptotic behavior is correct above and below the critical speed.
